# Solutions for Chapter 4

#### 4E1{-}
In the model definition below, which line is the likelihood?
$$\rightarrow ~ y_i \sim Normal(\mu, \sigma) ~ \leftarrow\\
\mu \sim Normal(0, 10) \\
\sigma \sim Exponential(1)$$

#### 4E2{-} 
In the model definition just above, how many parameters are in the posterior distribution?

**Two: $\mu$ and $\sigma$.

#### 4E3{-}
Using the model definition above, write down the appropriate form of Bayes’ theorem that includes the proper likelihood and priors.
$$Pr(\mu, \sigma | y) = \frac{\prod_i Normal(y_i| \mu, \sigma) \cdot Normal(mu | 0, 10) \cdot Exponential(\sigma | 1)}
{\int \int \prod_i Normal(y_i| \mu, \sigma) \cdot Normal(mu | 0, 10) \cdot Exponential(\sigma | 1) \partial \mu \partial \sigma}

$$


#### 4M1{-}
For the model definition below, simulate observed y values from the prior (not the posterior).
$$y_i \sim Normal(\mu,\sigma) \\
\mu \sim Normal(0, 10) \\
\sigma \sim Exponential(1)$$
```{r 4M1}
library(ggplot2)

samples_N <- 1e4
mu <- rnorm(samples_N, 0, 10)
sigma <- rexp(samples_N, 1)
y_sim <- rnorm(samples_N, mu, sigma)

ggplot(data=NULL, aes(x=y_sim)) + 
  geom_histogram(bins=50) + 
  xlab("Simulated Y")
```

#### 4M2 {-}
Translate the model just above into a quap formula.

```{r 4M2}
quap_formula <- alist(
  y ~ dnorm(mu, sigma),
  mu ~ dnorm(0,10),
  sigma ~ dexp(1))
```

