# Ordered Categorical Data, i.e., Likert-scales


```{r echo=FALSE, warning=FALSE, message=FALSE}
library(rethinking)
library(tidyverse)
```

One of a very popular type of response in psychology and social sciences are Likert-scale response. For example, you may be asked to respond on how attractive you find a person in a photo from 1 (very unattractive) to 7 (very attractive). Or to respond how satisfied you are with a service from 1 (very unsatisfied) to 4 (very satisfied). Or rate your confidence on a 5-point scale, etc. Likert-scale responses are extremely common and are quite often analyzed via linear models (i.e., _t_-test, repeated measures ANOVA, linear-mixed models). The purpose of these notes is to document both technical and, more importantly, conceptual problems this approach entails. 

## Conceptualization of responses: continuous variable discritized via a set of cut-points
First, let us think what responses correspond to as it will become very important once we discuss conceptual problems with using linear models for Likert-scale data. 

When we ask a participant to respond "On scale from 1 to 7, how attractive do you find the face in the photo?", we assume that there is a _continuous_ internal variable (encoded via a neural ensemble) that represents attractiveness of a face (or our satisfaction with service, or our confidence, etc.). The strength of that representation varies in a continuous manner from its minimum (e.g., baseline firing rate, if we assume that strength is encoded by spiking rate) to maximum (maximum firing rate for that neural ensemble). When we impose a seven-point scale on a participants, we force them to discretize this continuous variable, creating a _many-to-one_ mapping. In other words, a participant decides that values (intensities) within a particular range all get mapped on _1_, a different but adjacent range of higher values corresponds to _2_, etc. You can think about it as values within that range being "rounded" (regressed?) towards the mean. Or, equivalently, you can think in terms of cut-points that defined range limits. This is how the discretization is depicted in the figure below. In addition, our participant has a set of cut-points that discretize this signal, as depicted in the figure below. If the signal is below the first cut point, our participant response is "1". When it is between first and second cut points, the response is "2" and so on. When it is to the right of the last sixth cut point, it is "7". This conceptualization means that responses are an ordered categorical variable, as any underlying intensity for a response "1" is necessarily smaller than _any_ intensity for response "2" and both are smaller than, again, _any_ intensity for response "3".

```{r echo=FALSE}
intensity <- c(0.5, 1.7, 6.4)
ggplot() +
  geom_hline(yintercept = 1:6) +
  geom_hline(yintercept = c(0, 7), color="red", size=1)+ 
  geom_bar(data = tibble(x=1:length(intensity), y=intensity), aes(x=x, y=y), stat="identity") +
  scale_y_continuous(name = "Signal intensity", breaks = c(0, 7), labels = c("Minimal signal", "Maximal signal"), limits = c(-0.5, 7.5),
                     sec.axis = dup_axis(name="Cut points", breaks = 1:6, labels = 1:6)) + 
  scale_x_continuous(name = "Intensity to Response", breaks = 1:length(intensity), labels = sprintf("%.1f→%d", intensity, ceiling(intensity))) +
  coord_flip() +
  labs(subtitle="Many-to-one mapping of a continuous variable on categorical responses")
```

As per usual, we assume that our continuous variable is noisy and its values can be described as being drawn from a normal distribution centered at the "true" intensity level^[Here, I will assume that cut-points are fixed and it is parameters of the normal distribution that get adjusted. The actual implementation of ordered categorical has it other way around, so that intensity always comes from a normal distribution centered at 0 and with a standard deviation of one and it is cut-points that get adjusted. The two are mathematically equivalent but I find the former to be more intuitive.]. Here, the consistency of responses will depend on the width (standard deviation) of this distribution. The broader this distribution and / or closer it is to a cut-point, the more variable discrete responses will be.

```{r echo=FALSE}
x <- seq(0, 7, length.out = 500)
gaussians <- bind_rows(
  tibble(x = x, label = "1") %>%
    mutate(y = dnorm(x, mean = 2.5, sd = 0.125)),
    tibble(x = x, label = "2") %>%
    mutate(y = dnorm(x, mean = 2.5, sd = 1)),
    tibble(x = x, label = "3") %>%
    mutate(y = dnorm(x, mean = 2, sd = 0.125))
  ) %>%
  mutate(Response = ifelse(x == 0, 1, ceiling(x)),
         Response = ifelse(Response > 7, 7, Response),
         Response = factor(Response, levels = 1:7))

ggplot() +
  geom_ribbon(data=gaussians, aes(x=x, ymax=y, fill=Response), ymin=0) +
  geom_vline(xintercept = 1:6) +
  geom_vline(xintercept = c(0, 7), color="red", size=1)+ 
  scale_x_continuous(name = "Signal intensity", breaks = c(0, 7), labels = c("Minimal signal", "Maximal signal"), limits = c(-0.5, 7.5),
                     sec.axis = dup_axis(name="Cut points", breaks = 1:6, labels = 1:6)) + 
  facet_wrap(label~., ncol=1) + 
  theme(strip.background = element_blank(),  strip.text.x = element_blank()) +
  scale_y_continuous(name = "", breaks = NULL)

```

Our goal is to recover both the cut-points and the normal distribution using only observed responses.


## Conceptual problem with linear models: we change our mind about what responses correspond to.
If we use a linear model (a _t_-test, a repeated-measures ANOVA, linear-mixed models, etc.) to fit responses as an outcome (dependent) variable, we change the narrative. Recall that our original (and very intuitive) conceptualization was that they reflect a many-to-one mapping between an underlying continuous variable and a discrete (ordered categorical) response. Instead, linear models assume a _one-to-one_ mapping between the internal variable and observed responses. For example, a linear model postulates that for a 7-point Likert scale participant _could have_ responded with 6.5, 3.14, or 2.71828 but, for whatever reason (sheer luck?), we only observed a handful of integer values. 

Notice that this is _not_ how we think participants behave. I think everyone^[Never say always!] would object to the idea that the limited repertoire of responses is due to endogenous processing rather than exogenous limitations imposed by an experimental design. Yet, this is how a _linear model_ thinks about it and, if you are not careful, it is easy to miss this change in the narrative. It is, however, important as it means that estimates produced by a linear model are about that alternative one-to-one kind of responses, not the many-to-one that you had in mind! That alternative is not a bad story per se, it is just a _different_ story that should not be confused with the original one.

This change in the narrative of what responses correspond to is also a problem if you want to use a (fitted) linear model to simulate the data. It will happily spit out real valued responses like 6.5, 3.14, or 2.71828 (if you feel lucky enough to expect an actual integer response, you'd better use this luck on playing an actual lottery). You have two options. First, you bite the bullet and take them at their face value, sticking to one-to-one mapping between an internal variable and an observed response. That let's you keep the narrative but means that real and ideal observers play by different rules. Their responses are different and that means your conclusions based on an ideal observer behavior are of limited use. Alternatively, you can round real-valued responses off to a closest integer getting categorical responses. Unfortunately, that means changing the narrative yet again. In this case, you fitted the model assuming one-to-one mapping but you use its predictions assuming many-to-one. Not good. It is really hard to understand what is going on, if you keep changing your mind on what a response means. A linear model will also generate out-of-range responses, like -1 or 8. Here, you have little choice but to clip them into the valid range, forcing the many-to-one mapping on at least some responses. Again, change of a narrative means that model fitting and model interpretation rely on different conceptualizations of what response is.

This may sound too conceptual but I suspect that few people, who use linear models on Likert-scale data, realize that their model is not doing what they think it is doing and, erroneously!, interpret one-to-one linear-model estimates as many-to-one. The difference may or may not be crucial but the question is: Why employ a model that does something different to what you need to begin with? Remember, using an appropriate model and interpreting it correctly is _your_ job, not that of a mathematical model or nor is it a job of a software package. 

## A technical problem: Data that bunches up near a range limit.
When you use a linear model, you assume that residuals are normally distributed. This is something that you may not be sure of _before_ you fit a specific model, as it is residuals not the data that must be normally distributed. However, in some cases you may be fairly certain that this will not be the case, such as when a variable has only a limited range of values and the mean (the model prediction) is close to one of these limits. Whenever you have observations that are close to that hard limit, they will "bunch up" against it because they cannot go lower or higher than that. See the figure below for an illustration of how it happens if a _continuous_ variable $x$ is restricted to 1 to 7 range^[Note that for skewed distributions their mode is different from the mean!].
```{r echo=FALSE}
x <- seq(0, 1, length.out = 100)
mu <- c(2, 4, 5)
curves_df <- purrr::map_dfr(mu, ~tibble(y = rethinking::dbeta2(x, (. - 1)/ 6, 10), x = x * 6 + 1, `μ`= as.character(.)))

ggplot(data = curves_df, aes(x = x, y = y, color = `μ`)) + 
  geom_line() +
  ylab("Probability density") +
  scale_x_continuous("Continuous variable with 1..7 range", breaks = 1:7)
```

Note that the presence of a limit is not a deal breaker for using linear models per se. Most physical measures cannot be negative^[Try coming up with one and don't say "temperature"!] but as long your observations are sufficiently far away from it, you are fine. You cannot have a negative height but you certainly can use linear models for adult height as, for example, an average female height in USA 164±6.4 cm. In other words, the mean is more than 25 standards deviations away from the range limit of zero and the latter can be safely ignored.

Unfortunately, Likert-scale data combines an extremely limited range with a very coarse step. Even a 7-point Likert scale does not give you much of a wiggle room and routinely used 5-point scales are even narrower. This means that unless the mean is smack in the middle (e.g., at four for a 7-point scale), you are getting closer to one of the limits and your residuals become either positively (when approaching a lower limit) or negatively (for the upper one) skewed. In other words, the residuals are _systematically_ not normally distributed and their distributions depends on the mean. This clearly violates an assumption of normality of residuals and of their conditional i.i.d. (Independent and Identically Distributed). This is a deal breaker for parametric frequentist statistics (a _t_-test, a repeated-measures ANOVA, linear-mixed models), so that inferences are unreliable and are most probably not to be trusted.

## Another technical problem: Can we assume that cut-points are placed at equal intervals?
The skewed residuals described are a fundamental problem for parametric frequentist methods but is not critical if you use Bayesian or non-parametric bootstrapping/permutation linear models. Does this mean it is safe to use them? Probably not because when you use responses directly, you assume that participants place cut-points for responses in an orderly fashion with equal intervals between them. If that is the case, you can assume, for example, that $(4 + 4) / 2$ is equal to $(3 + 5) / 2)$ to $(2 + 6) / 2$ to $(1 + 7)/ 2$. However, what if this is _not_ the case, what if the cut-points (responses) are not placed at equal intervals? Then our basic arithmetic stops working! Take a look at the figure below.



